Data Provenance Tracking for Concurrent Programs

### 2014-08-31 paper.tex
new format for CGO sub
revisions - onto 10.75 pgs

\documentclass[preprint,10pt]{sigplanconf}
\usepackage{url}
\usepackage{listings}
\usepackage{subfigure}
\usepackage{times}
\usepackage{multirow}
\usepackage{color}
\usepackage{multicol}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{setspace}
\usepackage{graphicx}
\usepackage{xspace}
\usepackage{color}
\usepackage{rotating}

\newcommand{\addtodo}[1]{\textcolor{red}{[To do: #1]}}
\newcommand{\ctrap}{CTrap\xspace}
\newcommand{\ctraps}{CTraps\xspace}
\newcommand{\ctrapsfull}{CTraps-Full\xspace}
\newcommand{\ctrapsmm}{CTraps-NRR\xspace}
\newcommand{\lws}{CTraps-LWS\xspace}
\newcommand{\lwt}{LWT\xspace}

\newcommand{\Caption}[1]{\begin{minipage}{.95\columnwidth} \caption{#1} \end{minipage} \vspace{-1.2ex}}
\newcommand{\CaptionWide}[1]{\caption{#1}\vspace{-2ex}}
%\doublespacing
\authorinfo{Anonymous for Submission}{}

\title{Data Provenance Tracking for Concurrent Programs}

@@

%\author{Anonymous for Submission}
\begin{document}
\maketitle

\begin{abstract}
%Monitoring inter-thread memory dependences in production is useful.  Doing so
%improves debugging information and provides information to dynamic analyses and
%runtime systems.  Prior work that continuously monitors cross-thread
%dependences typically has overheads too high for production.  
We propose {\em last writer slicing}, a mechanism for tracking {\em data provenance information} in multithreaded code in a production setting.
Last writer slices dynamically track provenance of values by recording the thread and operation that last wrote each variable. We show that this information complements core dumps and greatly improves debugability. We also propose {\em communication traps}
(\ctraps), an extensible framework that uses last writer slices to interpose on operations that lead to inter-thread communication. We show how to use 
\ctraps implement multiple dynamic analysis tools for mulithreaded programs. Our evaluation on server programs and PARSEC benchmarks shows that last writer slices has low run time overheads (0--15\%) for many applications,including  
memcached, LevelDB, MySQL, and Apache. Our debugging case studies of
real, buggy programs shows that last writer slices are indeed useful for debugging.  
\end{abstract}

\section{Introduction}
Multi-threaded programming is challenging.  In contrast to simple sequential
reasoning, multi-threaded programs require complex reasoning about many threads
of control and their interactions.  In shared memory programs, threads interact
by reading and writing memory locations.  Threads' reads and writes interleave
arbitrarily and nondeterministically. Each interleaving can lead to different
behaviors and memory states, some of which may be undesirable, like crashes or
data corruption.  Moreover, because complete concurrency testing is infeasiable
in general, undesirable interleavings can evade testing and lead to production
failures.

The problem we are addressing in this work is that debugging support for
understanding thread interactions is inadequate.  Today, programmers use a
debugger to examine memory contents either during a debugging execution or by
loading a core dump from a failed execution.  Examining the contents of memory
reveals {\em what} the state of the program is ({\em e.g.}, the values of
variables).  Unfortunately, the real question programmers must answer is {\em
why} the state of the program is what it is.  Understanding why a program
entered an undesirable state ({\em e.g.}, crashed) is the key to changing the
code to prevent future executions from entering that state.  Multi-threading
makes this understanding elusive because implicating particular operations in
particular threads for a failure is a complex task.  Aiding developers in this
understanding for failed production executions is important because
hard-to-find bugs may manifest in production only.

@@

%During debugging provenance helps a programmer better understand not just where failures occur, but where values came from, connectng disparate
%parts of a program that interact, leading to a failure.

%In
%summary, the challenges in using provenance to debug concurrent programs are:
%(1) collecting the right information to characterize the concurrent execution
%behavior; and (2) making it efficient enough for general programs running in
%production settings.

@@

Motivated by these prior {\em ad hoc} efforts, we develop a dynamic analysis
framework called {\em Communication Traps} or {\em \ctraps}.  \ctraps runs
alongside a program and uses the provenance information provided by last writer
slices to identify memory accesses that lead to {\em communication} between the
threads. Communcation occurs when a thread accesses memory last written by a
different thread and \ctraps delivers a {\em trap} to a thread when it
communicates with another thread, giving it information about how communication
occurred.  Programmer-defined {\em trap handlers} can implement arbitrary
analysis using that information. By default, \ctraps has low overheads that are
appropriate production in many cases.  As we show in our evaluation, the
overhead of \ctraps applications scales with the complexity of their analysis.
Figure~\ref{fig:overview} illustrates the relationship between last writer
slices, \ctraps, and \ctraps applications.

\begin{figure}[h]
\centering
\includegraphics[width=.70\columnwidth]{figs/overview.pdf}
\Caption{\label{fig:overview}{\bf An overview of last writer slicing and CTraps.} Last writer slicing is a basic mechanism for tracking provenance.  The CTraps framework uses last writer slicing to identify communication and expose it to CTraps applications.  CTraps applications implement arbitrary dynamic concurrency analyses, like communication graph collection~\cite{recon} or cooperative debugging~\cite{cci}.}
\end{figure}

@@

\item{We describe last writer slicing, a novel data provenance tracking
mechanism, that produces information that is useful for debugging concurrent
programs.  We design and implement system support for collecting last writer
slices that is efficient enough for use in production.  }

\item{Building on last writer slicing, we develop \ctraps, an extensible
framework for implementing analyses of inter-thread communication in
multi-threaded programs.  }

\item{We evaluate the utility of last writer slices and \ctraps. We first
present a study of several real concurrent program bugs and show that last
writer slices are useful for debugging.  We then show how \ctraps is useful for
building dependence analyses by implementing two analyses from prior
work~\cite{cci,defuse,recon}. }

\item{We evaluate our system's performance using a variety of important
applications ({\em e.g.}, MySQL, memcached and the PARSEC
benchmarks~\cite{parsec}).  We show that in many cases our designs have
overheads low enough for use in deployment (0--15\%).}

@@

In this section we discuss debugging with data dependence information, using an
example, in order to motivate collecting last writer slices in deployed
systems.  We then provide context for \ctraps by discussing some more
sophisticated concurrency analyses that rely on last writer information.

@@

Prior work shows that some forms of data provenance information is directly
useful during program debugging.  The WhyLine
Debugger~\cite{whylinechi,whylineicse} showed that when programmers debug
programs, they benefit from the ability to directly ask such provenance
(``why'') questions of their debugger.  Bad Value Origin
Tracking~\cite{badapples} is a limited form of provenance analysis that tracks
intructions that stored certain unusable values.  The authors of Bad Value Origin Tracking showed that even such limited provenance information
helps understand some bug root causes. Like these
prior efforts, we focus on using provenance analysis for
{\em debugging} -- understanding the root cause of a failure -- as opposed to
{\em bug detection} -- finding new, unknown bugs.

@@

\subsection{\lwt Formalism and Correctness}
\label{sec:lwssoundness}
To discuss the correctness of our \lwt-based last writer slice support, we
first develop a formal description of last writer slices.  We consider an
N-threaded program $P = \{P_1, P_2, \ldots P_N\}$, where $P_t = (p^{t}_{1},
p^{t}_{2}, \ldots, p^{t}_{k_{t}})$ is the sequence of $k_{t}$ dynamic
instructions executed by thread $t$.  Without loss of generality, we assume all
instructions are memory reads and writes, with the form $Op(m)$ where $Op$ is
$Read$ or $Write$ and $m$ is a uniquely addressed memory location.
A point in an execution of $P$ after $e$ instructions have
executed is defined by $Q = (q_{0}, \ldots, q_{e})$ where each $q_{i} \in Q$ is a
$p^{t}_{j}$ from one of the $k_t$ instructions of thread $t$.  We introduce
$LWS[m]$ -- the Last Writer Slice for memory location $m$.   At any point $Q$,
$LWS[m]$ holds $(t,p^{t}_{\ell w})$, where $p^{t}_{\ell w} = max_{\ell}( 
q_{\ell} = Write(m) \in Q)$ and $t$ is the thread that
executed $p^{t}_{\ell w}$. Informally, $p^{t}_{\ell w}$ is the last instruction
in $Q$ to write $m$ and $t$ is the thread that executed it.  For an
unwritten location $n$, $LWS[n] = (\emptyset,\emptyset)$.

For our design to be correct, $LWT[m]$, the entry for a memory location $m$ in
our \lwt implementation, must be the same as $LWS[m]$ at all points in an
execution.  Correctness in the \lwt requires {\em soundness} and {\em
completeness}.  A {\em sound} design updates $LWT[m]$ only when $m$ is written.
A {\em complete} design always updates $LWT[m]$ when $m$ is written.  Our
compiler inserts \lwt updates before every write operation (for completeness)
and only before write operations (for soundness).  Formally, given a program,
$P$, our compiler pass transforms it to a new program, $P'$.  Each thread in
$P'$ executes a modified sequence of instructions: $P'_{t} \in P' = ( 
I(p^{t}_{1}), p^{t}_{1}, \ldots, I(p^{t}_{k_{t}}), p^{t}_{k_{t}} )$.
$I(p^{t}_{j})$ sets $LWT[m] = (t,p^{t}_{j})$ if and only if $p^{t}_{j} =
Write(m)$ and is a no-op otherwise.  This reasoning shows our \lwt is both
sound, complete, and thus correct.

Next we discuss two nuances of our design: (1) the interaction of our \lwt
implementation with data-races; and (2) the correctness of the
\lwt for accesses to overlapping memory regions.

\paragraph{Consistency and Atomicity}

Our implementation of the \lwt does not add any synchronization to the program.
As a result our analysis does not incur the overheads or possible thread
schedule perturbation synchronization can cause.  However, there are two
potential issues that stem from this design choice. First, we do not enforce
memory consistency of meta-data reads and writes in different threads.  Second,
our implementation does not enforce the atomicity of \lwt updates and their
corresponding program accesses.  In the absence of data-races our \lwt
implementation is {\em always correct} and neither of these issues exist.  Both
atomicity and ordering is enforced by the program synchronization that ensures
consistency for program data.

We use our \lwt formalism to show that without data-races, \lwt updates and
their corresponding memory accesses are atomic and well-ordered.  The relation
$p \prec q$ denotes that $p$ happens before $q$.  In correctly synchronized
code, any write by thread $t$ to $m$, $p^{t}_{j} = Write(m)$, is ordered by
program synchronization with $p^{t'}_{j'} = Op(m)$, a later access to $m$ by
any other thread $t'$ -- {\em i.e.}, $p^{t}_{j} \prec p^{t'}_{j'}$.  Our
compiler inserts an \lwt update $I(p^{t}_{j})$ into $P'$ before its write
operation, $p^{t}_{j}$. Thus, $I(p^{t}_{j}) \prec p^{t}_{j}$ due to program
order.  $I(p^{t'}_{j'})$ is inserted before $p^{t'}_{j'}$, but after the synchronization that orders $p^{t'}_{j'}$ and $p^{t}_{j}$, so $p^{t}_{j} \prec I(p^{t'}_{j'})$.  Happens-before is transitive, so without data-races, $I(p^{t}_{j})
\prec p^{t}_{j} \prec I(p^{t'}_{j'}) \prec p^{t'}_{j'}$, meaning $I(p^{t}_{j})$
happens before all accesses that $p^{t}_{j}$ happens before {\em and before
their respective \lwt updates}, $I(p^{t'}_{j'})$.  Symmetrically, for any
access $p^{t''}_{j''}$ in another thread $t''$ that precedes $p^{t}_{j}$,
$p^{t''}_{j''} \prec p^{t}_{j}$.  Our compiler inserts each $I(p^{t}_{j})$
immediately before each $p^{t}_{j}$, so $I(p^{t''}_{j''}) \prec p^{t''}_{j''}
\prec I(p^{t}_{j}) \prec p^{t}_{j}$, happens-before ordering $I(p^{t}_{j})$
after all preceding accesses to $m$ in other threads, {\em and after their
repective \lwt updates}.  Thus, at all points $Q$ in a data-race free
execution, given $q_r = I(p^{t}_{j})$, $q_p = p^{t}_{j}$, and $q_r \prec q_p$,
$\forall_{ \{q_s | q_r \prec q_s \prec q_p \}}: q_s \ne Op(m) \wedge q_s \ne
I(Op(m)) $.  In other words, \lwt updates and their accesses are atomic.

In a program with a data-race, there are some accesses, $p^{t}_{j}$ and
$p^{t'}_{j'}$, that are concurrent.  The \lwt updates for these accesses are
also concurrent because they are not transitively ordered by the accesses' ordering.  The unordered accesses and \lwt updates can interleave arbitrarily
because they are concurrent.  Some racy thread schedules may violate the
atomicity of accesses and their \lwt updates.  Such an execution could reach a
state $Q_{bad} = (\ldots, I(p^{t}_{j}), I(p^{t'}_{j'}), p^{t'}_{j'}, p^{t}_{j},
\ldots)$ and in that state, $LWS[m] = (t,p^{t}_{j})$, but $LWT[m] =
(t',p^{t'}_{j'})$.  The \lwt was effectively not updated when $p^{t}_{j}$
executed (violating completeness), or was effectively updated to reflect that
$p^{t'}_{j'}$ was $m$'s last writer, when $p^{t}_{j}$ was (violating
soundness).

According to the C++11 language specification, in the presence of a data-race,
a program's behavior is undefined and arbitrary state corruption can occur.  In
the presence of data-races, {\em any} analysis, including our \lwt analysis,
running on such a program could experience arbitrary corruption.  In practice
({\em e.g.}, on x86) these atomicity violations may cause limited \lwt
corruption, but even this is unlikely.  These atomicity violations only occur
only when accesses to the same data occur within a very short time interval,
which is uncommon, {\em and} an unserializable interleaving of accesses and
\lwt updates also occurs.

\paragraph{Non-overlapping Memory References} 
We assume written memory locations do not overlap in memory because our \lwt
design is {\em granularity oblivious}.  Each \lwt entry corresponds to a
location that was written.  Two instructions might reference different
locations that overlap because the referenced locations have different
granularities ({\em e.g.}, accessing a byte that is part of a previously
accessed word).  When that happens, both accesses update different \lwt
entries, because they reference different locations.  However, the overlapping
region then has two different entries reflecting the last {\em two} updates to
its parts rather than just its last write.

The \lwt remains correct, assuming the program only accesses non-overlapping
regions.  In programs that access overlapping regions, there is always some
entry in the \lwt that correctly reflects each memory location's last writer.
However, the coarse region's entry may be stale because of an
access to an overlapping region.

One approach to mitigating this issue would be to allocate \lwt entries for
regions of fixed granularity.  We chose not to do so for two reasons.  First,
associating \lwt entries with regions of  large granularity only ({\em i.e.},
cache-line) precludes sound analysis for fine-grained accesses.  Second,
associating \lwt entries with regions of small granularity ({\em i.e.}, byte)
requires multiple \lwt updates for larger granularity accesses, increasing
their overheads proportionately to the granularity of the access.

@@

\ctraps is a framework for analyzing and interposing on inter-thread
communication events during a program's execution.  We build \ctraps on top of
last writer slicing that we described in Section~\ref{sec:lastwriterslices}.
Using last writer slice information, \ctraps dynamically tracks 
inter-thread communication events. \ctraps exposes communication
events as {\em traps} during an execution that {\em trap handlers} can handle
to perform analysis and interposition.

@@

\subsection{Characterizing \ctraps}
\label{sec:char}
We characterized the instrumentation pruned by escape analysis.  We report our
results in Table~\ref{tab:char}.  The table shows the percent of
calls pruned and the absolute number of sites in parenthesis.

The data show that escape analysis can prove a substantial fraction of
operations are local and do not require instrumentation.  Across PARSEC,
between 8\% and 30\% of accesses were proved local.  In the server programs, a
larger proportion of accesses were able to be proved local (21\% -- 66\%).  The
difference is likely due to the fact that server threads perform more
independent work (as discussed in Section~\ref{sec:eval:parsecserver}).
Eliminating instrumentation using escape analysis is key to high performance.

\begin{table}[htb]
\centering
\small
\begin{tabular}{l | r r }
\multirow{3}{*}{\bf Application} & \multicolumn{2}{c}{\bf \em Pruned } \\
               & \multicolumn{2}{c}{\bf \em Static Local}  \\ 
               & \multicolumn{2}{c}{\bf \em Operations}    \\ \hline

Apache         &  21.4\%&(12K)                    \\
MySQL          &  21.3\%&(53K)                    \\
memcached      &  65.8\%&(7.5K)                   \\
LevelDB        &  44.0\%&(4.5K)                   \\ \hline
blackscholes   &   7.5\%&(4)                      \\
dedup          &   22.0\%&(171)                   \\
canneal        &   28.6\%&(226)                   \\
x264           &   29.8\%&(5960)                  \\
vips           &   24.6\%&(9116)                  \\
ferret         &   20.9\%&(792)                   \\
fluidanimate   &   19.6\%&(149)                   \\
streamcluster  &   11.6\%&(47)                    \\
swaptions      &   8.3\% &(19)                    \\
\end{tabular}
\Caption{\label{tab:char}Instrumentation pruned by static analysis.}
\end{table}

\section{Related Work}

>>>>

\newcount\draft\draft=1 % set to 0 for submission or publication
\newcount\cameraready\cameraready=0

\documentclass[pageno,nohyperref]{jpaper}

\ifnum\draft=1
  \input{revision}
  \usepackage{drafthead}
\fi
\usepackage{xxx}

\usepackage{graphicx}
\DeclareGraphicsExtensions{.pdf,.jpg,.png}
\graphicspath{{./figs/} {./plots/}}

\usepackage[normalem]{ulem}

\usepackage{listings}
\lstset{
    language=C,
    emphstyle=\bfseries,
    basicstyle=\ttfamily\small,
    aboveskip=1mm plus 1mm minus 1mm,
    belowskip=1mm plus 1mm minus 1mm,
    mathescape=true,
    xleftmargin=\parindent,
}
\newcommand{\lil}{\lstinline}

\usepackage[labelfont=bf,font=small]{caption}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathptmx}
\usepackage{multirow}
%\usepackage{multicol}
\usepackage{color}
\usepackage{rotating}

\usepackage{xspace}
\newcommand{\addtodo}[1]{\textcolor{red}{[To do: #1]}}
\newcommand{\ctrap}{CTrap\xspace}
\newcommand{\ctraps}{CTraps\xspace}
\newcommand{\ctrapsfull}{CTraps-Full\xspace}
\newcommand{\ctrapsmm}{CTraps-NRR\xspace}
\newcommand{\lws}{LWS\xspace}
\newcommand{\lwt}{LWT\xspace}

\newcommand{\Caption}[1]{\begin{minipage}{.95\columnwidth} \caption{#1} \end{minipage} \vspace{-1.2ex}}
\newcommand{\CaptionWide}[1]{\caption{#1}\vspace{-2ex}}
%\doublespacing
%\authorinfo{Anonymous for Submission}{}

@@

%\author{Anonymous for Submission}

\title{Data Provenance Tracking for Concurrent Programs}
\newcommand{\submissionnumber}{65}
\date{}

\begin{document}

\special{papersize=8.5in,11in}
\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}

\maketitle

\thispagestyle{empty}
\begin{abstract}
%Monitoring inter-thread memory dependences in production is useful.  Doing so
%improves debugging information and provides information to dynamic analyses and
%runtime systems.  Prior work that continuously monitors cross-thread
%dependences typically has overheads too high for production.  
We propose {\em last writer slicing}, a mechanism for tracking {\em data
provenance information} in multithreaded code in a production setting.  Last
writer slices dynamically track provenance of values by recording the thread
and operation that last wrote each variable. We show that this information
complements core dumps and greatly improves debugability. We also propose {\em
communication traps} (\ctraps), an extensible framework that uses last writer
slices to interpose on operations that lead to inter-thread communication. We
show how to use \ctraps implement multiple dynamic analysis tools for
mulithreaded programs. Our evaluation on server programs and PARSEC benchmarks
shows that last writer slices has low run time overheads (0--15\%) for many
applications,including  memcached, LevelDB, MySQL, and Apache. Our debugging
case studies of real, buggy programs shows that last writer slices are indeed
useful for debugging.  
\end{abstract}

\section{Introduction}
Multi-threaded programming is challenging.  In contrast to simple sequential
reasoning, multi-threaded programs require complex reasoning about many threads
and their interactions.  Threads primarily interact by reading and writing
shared memory locations and different threads' reads and writes interleave
arbitrarily and nondeterministically. Different interleavings can lead to
different behaviors, some of which may be undesirable, like crashes or data
corruption. Such undesirable interleavings often evade testing and cause
production failures because complete concurrency testing is infeasible.

The problem that this work addresses is that debugging support for
understanding thread interactions is inadequate.  Today, programmers use a
debugger to examine memory state during a debugging execution or by loading a
core dump after a failure.  Examining memory reveals {\em what} the state of
the program is ({\em e.g.}, ``is the pointer null?'').  Unfortunately, the real
question programmers must answer is {\em why} the state of the program is what
it is ({\em e.g.}, ``what thread and instruction made the pointer null?'').
Understanding why a program entered an undesirable state ({\em e.g.}, crashed)
is the key to fixing buggy code to prevent that state.  Aiding developers in
this understanding for failed production executions is especially important
because hard-to-find bugs may manifest in production only.

@@

@@

Motivated by these prior {\em ad hoc} efforts, we develop a dynamic analysis
framework called {\em Communication Traps} or {\em \ctraps}.  \ctraps runs
alongside a program and uses the provenance information provided by last writer
slices to identify memory accesses that lead to {\em communication} between the
threads. Communcation occurs when a thread accesses memory last written by a
different thread and \ctraps delivers a software {\em trap} to a thread when it
communicates with another thread, giving it information about how communication
occurred.  Programmer-defined {\em \ctraps Applications} can implement arbitrary
analysis using that information. By default, \ctraps has low overheads, 
appropriate for production in many cases.  As our evaluation shows, the
overhead of \ctraps applications scales with the complexity of their analysis.
Figure~\ref{fig:overview} illustrates the relationship between last writer
slices, \ctraps, and \ctraps applications.

\begin{figure}[h]
\centering
\includegraphics[width=\columnwidth]{figs/systemoverview.pdf}
\Caption{\label{fig:overview}{\bf System Overview.} Last writer slicing tracks provenance, which augments the core dump and enables provenance debugging.  CTraps uses provenance to convey communication to CTraps applications, which implement concurrency analyses, like communication graphs~\cite{recon} or CBI~\cite{cci}.}
\end{figure}

@@

\item{{\bf Last writer slicing} ({\em LWS}): a provenance tracking mechanism
useful for concurrency debugging.  We present an \lws implementation
efficient enough for production use.}

\item{{\bf Communication traps} (\ctraps): an extensible
framework for implementing concurrency analyses, built on \lws.  }

\item{{\bf Efficacy evaluation}: We show that \lws is useful for debugging with
a study of several real concurrency bugs.  We show \ctraps is useful for
building concurrency analyses by implementing two analyses from prior
work~\cite{cci,defuse,recon}. }

\item{{\bf Performance evaluation}: We show that in many cases our designs have
overheads low enough for use in deployment (0--15\%) with a study of a diverse
set of real programs ({\em e.g.}, MySQL, memcached and the PARSEC
benchmarks~\cite{parsec}).  }

@@

In this section we discuss debugging with data dependence information, using an
example, in order to motivate collecting last writer slices in deployed
systems.  We then provide context for \ctraps by discussing more sophisticated
analyses that use provenance.

@@

Prior work shows that data provenance information is directly useful during
program debugging.  The WhyLine Debugger~\cite{whylinechi,whylineicse} showed
that when programmers debug, they benefit from the ability to ask provenance
(``why'') questions of their debugger.  Bad Value Origin
Tracking~\cite{badapples} is a limited form of provenance analysis that tracks
intructions that stored certain unusable values.  The authors of Bad Value
Origin Tracking showed that even such limited provenance information helps
understand some bug root causes. Like these prior efforts, we focus on using
provenance analysis for {\em debugging} -- understanding the root cause of a
failure -- as opposed to {\em bug detection} -- finding new, unknown bugs.

@@

\subsection{\lwt Formalism and Correctness}
\label{sec:lwssoundness}

We developed a formalism of our \lwt-based last writer slices support, but we
omit it due to space constraints.  We used our formalism to show that the \lwt
correctly records last writer slices.  There are two important caveats in our
formal discussion.  The first caveat is that the \lwt implementation introduces
no synchronization.  We proved that the \lwt is always correct in
data-race-free programs because \lwt accesses are synchronized by program
synchronization.  However, if an execution has a data-race, program memory
operations and \lwt accesses are no longer atomic.  If such a data-race leads
to an unserializable interleaving of \lwt accesses, the memory location's \lwt
entry may reflect the wrong last writer.  The second caveat is that the \lwt
assumes that a memory location accessed at one granularity ({\em e.g.},
byte-aligned) is never accessed at a different granularity ({\em e.g.}, as part
of a word-aligned access).  Assuming accesses do not overlap ensures there is
always only one \lwt entry for each memory location.

%To discuss the correctness of our \lwt-based last writer slice support, we
%first develop a formal description of last writer slices.  We consider an
%N-threaded program $P = \{P_1, P_2, \ldots P_N\}$, where $P_t = (p^{t}_{1},
%p^{t}_{2}, \ldots, p^{t}_{k_{t}})$ is the sequence of $k_{t}$ dynamic
%instructions executed by thread $t$.  Without loss of generality, we assume all
%instructions are memory reads and writes, with the form $Op(m)$ where $Op$ is
%$Read$ or $Write$ and $m$ is a uniquely addressed memory location.
%A point in an execution of $P$ after $e$ instructions have
%executed is defined by $Q = (q_{0}, \ldots, q_{e})$ where each $q_{i} \in Q$ is a
%$p^{t}_{j}$ from one of the $k_t$ instructions of thread $t$.  We introduce
%$LWS[m]$ -- the Last Writer Slice for memory location $m$.   At any point $Q$,
%$LWS[m]$ holds $(t,p^{t}_{\ell w})$, where $p^{t}_{\ell w} = max_{\ell}( 
%q_{\ell} = Write(m) \in Q)$ and $t$ is the thread that
%executed $p^{t}_{\ell w}$. Informally, $p^{t}_{\ell w}$ is the last instruction
%in $Q$ to write $m$ and $t$ is the thread that executed it.  For an
%unwritten location $n$, $LWS[n] = (\emptyset,\emptyset)$.
%
%For our design to be correct, $LWT[m]$, the entry for a memory location $m$ in
%our \lwt implementation, must be the same as $LWS[m]$ at all points in an
%execution.  Correctness in the \lwt requires {\em soundness} and {\em
%completeness}.  A {\em sound} design updates $LWT[m]$ only when $m$ is written.
%A {\em complete} design always updates $LWT[m]$ when $m$ is written.  Our
%compiler inserts \lwt updates before every write operation (for completeness)
%and only before write operations (for soundness).  Formally, given a program,
%$P$, our compiler pass transforms it to a new program, $P'$.  Each thread in
%$P'$ executes a modified sequence of instructions: $P'_{t} \in P' = ( 
%I(p^{t}_{1}), p^{t}_{1}, \ldots, I(p^{t}_{k_{t}}), p^{t}_{k_{t}} )$.
%$I(p^{t}_{j})$ sets $LWT[m] = (t,p^{t}_{j})$ if and only if $p^{t}_{j} =
%Write(m)$ and is a no-op otherwise.  This reasoning shows our \lwt is both
%sound, complete, and thus correct.
%
%Next we discuss two nuances of our design: (1) the interaction of our \lwt
%implementation with data-races; and (2) the correctness of the
%\lwt for accesses to overlapping memory regions. 
%
%\paragraph{Consistency and Atomicity}
%
%Our implementation of the \lwt does not add any synchronization to the program.
%As a result our analysis does not incur the overheads or possible thread
%schedule perturbation synchronization can cause.  However, there are two
%potential issues that stem from this design choice. First, we do not enforce
%memory consistency of meta-data reads and writes in different threads.  Second,
%our implementation does not enforce the atomicity of \lwt updates and their
%corresponding program accesses.  In the absence of data-races our \lwt
%implementation is {\em always correct} and neither of these issues exist.  Both
%atomicity and ordering is enforced by the program synchronization that ensures
%consistency for program data.   
%
%We use our \lwt formalism to show that without data-races, \lwt updates and
%their corresponding memory accesses are atomic and well-ordered.  The relation
%$p \prec q$ denotes that $p$ happens before $q$.  In correctly synchronized
%code, any write by thread $t$ to $m$, $p^{t}_{j} = Write(m)$, is ordered by
%program synchronization with $p^{t'}_{j'} = Op(m)$, a later access to $m$ by
%any other thread $t'$ -- {\em i.e.}, $p^{t}_{j} \prec p^{t'}_{j'}$.  Our
%compiler inserts an \lwt update $I(p^{t}_{j})$ into $P'$ before its write
%operation, $p^{t}_{j}$. Thus, $I(p^{t}_{j}) \prec p^{t}_{j}$ due to program
%order.  $I(p^{t'}_{j'})$ is inserted before $p^{t'}_{j'}$, but after the synchronization that orders $p^{t'}_{j'}$ and $p^{t}_{j}$, so $p^{t}_{j} \prec I(p^{t'}_{j'})$.  Happens-before is transitive, so without data-races, $I(p^{t}_{j})
%\prec p^{t}_{j} \prec I(p^{t'}_{j'}) \prec p^{t'}_{j'}$, meaning $I(p^{t}_{j})$
%happens before all accesses that $p^{t}_{j}$ happens before {\em and before
%their respective \lwt updates}, $I(p^{t'}_{j'})$.  Symmetrically, for any
%access $p^{t''}_{j''}$ in another thread $t''$ that precedes $p^{t}_{j}$,
%$p^{t''}_{j''} \prec p^{t}_{j}$.  Our compiler inserts each $I(p^{t}_{j})$
%immediately before each $p^{t}_{j}$, so $I(p^{t''}_{j''}) \prec p^{t''}_{j''}
%\prec I(p^{t}_{j}) \prec p^{t}_{j}$, happens-before ordering $I(p^{t}_{j})$
%after all preceding accesses to $m$ in other threads, {\em and after their
%repective \lwt updates}.  Thus, at all points $Q$ in a data-race free
%execution, given $q_r = I(p^{t}_{j})$, $q_p = p^{t}_{j}$, and $q_r \prec q_p$,
%$\forall_{ \{q_s | q_r \prec q_s \prec q_p \}}: q_s \ne Op(m) \wedge q_s \ne
%I(Op(m)) $.  In other words, \lwt updates and their accesses are atomic.
%
%In a program with a data-race, there are some accesses, $p^{t}_{j}$ and
%$p^{t'}_{j'}$, that are concurrent.  The \lwt updates for these accesses are
%also concurrent because they are not transitively ordered by the accesses' ordering.  The unordered accesses and \lwt updates can interleave arbitrarily
%because they are concurrent.  Some racy thread schedules may violate the
%atomicity of accesses and their \lwt updates.  Such an execution could reach a
%state $Q_{bad} = (\ldots, I(p^{t}_{j}), I(p^{t'}_{j'}), p^{t'}_{j'}, p^{t}_{j},
%\ldots)$ and in that state, $LWS[m] = (t,p^{t}_{j})$, but $LWT[m] =
%(t',p^{t'}_{j'})$.  The \lwt was effectively not updated when $p^{t}_{j}$
%executed (violating completeness), or was effectively updated to reflect that
%$p^{t'}_{j'}$ was $m$'s last writer, when $p^{t}_{j}$ was (violating
%soundness).    
%
%According to the C++11 language specification, in the presence of a data-race,
%a program's behavior is undefined and arbitrary state corruption can occur.  In
%the presence of data-races, {\em any} analysis, including our \lwt analysis,
%running on such a program could experience arbitrary corruption.  In practice
%({\em e.g.}, on x86) these atomicity violations may cause limited \lwt
%corruption, but even this is unlikely.  These atomicity violations only occur
%only when accesses to the same data occur within a very short time interval,
%which is uncommon, {\em and} an unserializable interleaving of accesses and
%\lwt updates also occurs.
%
%\paragraph{Non-overlapping Memory References} 
%We assume written memory locations do not overlap in memory because our \lwt
%design is {\em granularity oblivious}.  Each \lwt entry corresponds to a
%location that was written.  Two instructions might reference different
%locations that overlap because the referenced locations have different
%granularities ({\em e.g.}, accessing a byte that is part of a previously
%accessed word).  When that happens, both accesses update different \lwt
%entries, because they reference different locations.  However, the overlapping
%region then has two different entries reflecting the last {\em two} updates to
%its parts rather than just its last write.
%
%The \lwt remains correct, assuming the program only accesses non-overlapping
%regions.  In programs that access overlapping regions, there is always some
%entry in the \lwt that correctly reflects each memory location's last writer.
%However, the coarse region's entry may be stale because of an
%access to an overlapping region.
%
%One approach to mitigating this issue would be to allocate \lwt entries for
%regions of fixed granularity.  We chose not to do so for two reasons.  First,
%associating \lwt entries with regions of  large granularity only ({\em i.e.},
%cache-line) precludes sound analysis for fine-grained accesses.  Second,
%associating \lwt entries with regions of small granularity ({\em i.e.}, byte)
%requires multiple \lwt updates for larger granularity accesses, increasing
%their overheads proportionately to the granularity of the access.

@@

\ctraps is a framework for analyzing and interposing on inter-thread
communication.  \ctraps uses \lws (\S~\ref{sec:lastwriterslices}) to track
inter-thread communication.  \ctraps exposes communication events as {\em
traps} during an execution that {\em trap handlers} can handle to perform
analysis and interposition.

@@

%\subsection{Characterizing \ctraps}
%\label{sec:char}
%We characterized the instrumentation pruned by escape analysis.  We report our
%results in Table~\ref{tab:char}.  The table shows the percent of
%calls pruned and the absolute number of sites in parenthesis.
%
%The data show that escape analysis can prove a substantial fraction of
%operations are local and do not require instrumentation.  Across PARSEC,
%between 8\% and 30\% of accesses were proved local.  In the server programs, a
%larger proportion of accesses were able to be proved local (21\% -- 66\%).  The
%difference is likely due to the fact that server threads perform more
%independent work (as discussed in Section~\ref{sec:eval:parsecserver}).
%Eliminating instrumentation using escape analysis is key to high performance.  
%
%\begin{table}[htb]
%\centering
%\small
%\begin{tabular}{l | r r }
%\multirow{3}{*}{\bf Application} & \multicolumn{2}{c}{\bf \em Pruned } \\
%               & \multicolumn{2}{c}{\bf \em Static Local}  \\ 
%               & \multicolumn{2}{c}{\bf \em Operations}    \\ \hline
%
%
%Apache         &  21.4\%&(12K)                    \\
%MySQL          &  21.3\%&(53K)                    \\
%memcached      &  65.8\%&(7.5K)                   \\
%LevelDB        &  44.0\%&(4.5K)                   \\ \hline
%blackscholes   &   7.5\%&(4)                      \\
%dedup          &   22.0\%&(171)                   \\
%canneal        &   28.6\%&(226)                   \\
%x264           &   29.8\%&(5960)                  \\
%vips           &   24.6\%&(9116)                  \\
%ferret         &   20.9\%&(792)                   \\
%fluidanimate   &   19.6\%&(149)                   \\
%streamcluster  &   11.6\%&(47)                    \\
%swaptions      &   8.3\% &(19)                    \\
%\end{tabular}
%\Caption{\label{tab:char}Instrumentation pruned by static analysis.}
%\end{table}
%
%
\section{Related Work}
